Spatio-temporal estimates of HIV risk group proportions for adolescent girls and young women across 13 priority countries in sub-Saharan Africa

### 2022-09-27 src/docs_28-09-22-toronto/28-09-22-toronto.Rmd
Iterate Toronto presentation forward

---
title: Spatio-temporal estimates of HIV risk group proportions for adolescent girls and young women across 13 priority countries in sub-Saharan Africa
subtitle: Centre for Global Health Research, UToronto
author: Adam Howes
institute: Imperial College London
date: May 2022
bibliography: citations.bib
header-includes:
  - \widowpenalties 1 150
output:
  beamer_presentation:
    latex_engine: pdflatex
    highlight: haddock
    fig_width: 7 
    fig_height: 3
    includes:
        in_header: preamble.tex
---

@@

* Prevention
  * Core package
  * Intensified interventions
* It's important to priorisite intensified interventions to those at highest risk

@@

The Global AIDS strategy 2021-2016 proposed stratifying HIV prevention for AGYW based upon

1. Population-level HIV incidence
2. Individual-level sexual risk behaviour

@@

1. Enable implementation of prevention stratified by incidence and behaviour
2. Assess the benefits of such approaches

@@

1. Estimate the proportion of AGYW in four behavioural risk groups at a district level (in 13 countries identified as priority by The Global Fund)
2. Analyze numbers of new infections reached by stratified prevention strategies

@@

* We used sexual behaviour data from AIS, BAIS, DHS and PHIA household surveys to place respondents into four risk groups:
  1. Not sexually active
  2. One cohabiting sexual partner
  3. Non-regular sexual partner(s)
  4. Female sex workers
* District-level HIV incidence, prevalence, population size estimates from the Naomi model [@eaton2021naomi]
* Risk ratios from ALPHA network analysis [@slaymaker2020croi] and UNAIDS analysis led by Keith Sabin

@@

### Stage 1: $k = 1, 2, 3^{+}$

* Multinomial logistic regression model for the proportion of AGYW in the $k = 1, 2, 3^{+}$ risk groups, using all 47 surveys
* Selected model (by CPO) included:
  * Age country effects (IID)
  * Country effects (IID)
  * Correlated spatial effects (ICAR)
  * Correlated temporal effects (AR1) 
* Multinomial-Poisson transformation allowed use of `R-INLA` for inference

### Stage 2: $k = 3, 4$

* Logistic regression model for the proportion of those in the $k = 3^{+} = \{3, 4\}$ risk groups who are in the $k = 4$ risk group, using only the 13 surveys with a specific transactional sex question
* Selected model (by CPO) included:
  * Age country effects (IID)
  * Country effects (IID)
  * Correlated spatial effects (ICAR)
  * Clients of FSW covariates [@hodgins2022population]
* Used `R-INLA` for inference

### Combination and FSW adjustment

* Take 1000 samples from each model, then multiply suitably to generate estimates for all four risk groups
* We adjusted the $k = 4$ risk group to match national FSW estimates from @johnston2022deriving

>>>>

---
title: Spatio-temporal estimates of HIV risk group proportions for adolescent girls and young women across 13 priority countries in sub-Saharan Africa
subtitle: Centre for Global Health Research, UToronto
author: Adam Howes
institute: Imperial College London
date: September 2022
bibliography: citations.bib
header-includes:
  - \widowpenalties 1 150
output:
  beamer_presentation:
    latex_engine: pdflatex
    highlight: haddock
    fig_width: 7 
    fig_height: 3
    includes:
        in_header: preamble.tex
---

@@

* Prevention options can be divided into two:
  1. Core package
  2. Intensified interventions
* There are not enough resources to offer the more costly intensified interventions to all AGYW, so it's important to prioritise those at highest risk

@@

* The Global AIDS strategy 2021-2016 proposed stratifying HIV prevention for AGYW based upon
  1. Population-level HIV incidence
  2. Individual-level sexual risk behaviour
* Takes into account the two most proximal drivers of sexual transmission

@@

1. Enable implementation of prevention stratified by incidence and behaviour
2. Assess the benefits of such approaches: is it worth it?

@@

1. Estimate the proportion of AGYW in four behavioural risk groups at a district level (in 13 countries identified as priority by The Global Fund)
2. Analyze the new infections which could be reached by different stratified prevention strategies

@@

* We used sexual behaviour data from AIS, BAIS, DHS and PHIA household surveys to place respondents into $K = 4$ risk groups:
  1. $k = 1$ Not sexually active
  2. $k = 2$ One cohabiting sexual partner
  3. $k = 3$ Non-regular sexual partner(s)
  4. $k = 4$ Female sex workers
* District-level HIV incidence, prevalence, population size estimates from the Naomi model [@eaton2021naomi]
* Risk ratios from ALPHA network analysis [@slaymaker2020croi] and UNAIDS analysis led by Keith Sabin

@@

* Only some of the surveys included a transactional sex question, required to differentiate between the $k = 3$ and $k = 4$ risk groups
* Our approach was to fit a two-stage model
  1. Spatio-temporal multinomial logistic regression model for the proportion of AGYW in the $k = 1, 2, 3^{+}$ risk groups, using all 47 surveys
  2. Spatial logistic regression model for the proportion of those in the $k = 3^{+} = \{3, 4\}$ risk groups who are in the $k = 4$ risk group, using only the 13 surveys with a specific transactional sex question

## Notation

* $k \in \{1, \ldots, 4\}$: risk groups
* $i \in \{1, \ldots, n\}$: districts
* $c[i] \in \{\text{Botswana}, \ldots, \text{Zimbabwe}\}$: country containing district $i$
* $t \in \{1999, \ldots, 2018\}$: years
* $a \in \{\text{15-19}, \text{20-24}, \text{25-29}\}$: age groups
* $y_{ita}^\star = (y_{ita1}^\star, y_{ita2}^\star, y_{ita3}^\star, y_{ita4}^\star)$: survey weighted multinomial observations
* $m_{ita}^\star$: survey weighted multinomial sample size

## Multinomial logistic regression model

* Would like to use integrated nested Laplace approximations for fast, accurate inference, but `R-INLA` is not compatible with multinomial likelihoods because they depend on multiple elements of the latent field
* Instead, use that multinomial logistic regression models can be recast as a Poisson log-linear models using the Poisson trick
* Include observation-specific random effects $\theta_{ita} \sim \mathcal{N}(0, 1000^2)$ in any linear predictor $\eta_{itak} = \theta_{ita} + \cdots$ to ensure exact reproduction of $m_{ita}^\star$

## Multinomial logistic regression model

* Consider models of the form
\begin{align}
    y_{itak}^\star &\sim \text{xPoisson}(\lambda_{itak}) \\
    \log(\lambda_{itak}) &= \theta_{ita} + \beta_k + \zeta_{c[i]k} + \alpha_{ac[i]k} + \phi_{ik} + \gamma_{tk} + \delta_{itk}.
\end{align}
* With the following random effects
  * $\theta_{ita}$: observation (IID)
  * $\beta_k$: category (IID)
  * $\zeta_{ck}$: country-category (IID x IID) 
  * $\alpha_{ack}$: age-country-category (IID x IID x IID)
  * $\phi_{ik}$: space-category ({IID, **Besag**} x IID)
  * $\gamma_{tk}$: year-category ({**IID**, AR1} x IID)
  * $\delta_{itk}$: space-year-category (Implemented but crashing on cluster at the moment...)

## Multinomial logistic regression model

* Independent penalised complexity [@simpson2017penalising] priors on all standard deviation parameters with $\sigma = 0$ and $\mathbb{P}(\sigma > 2.5 = 0.01)$
  * Sidenote, I'm interested as to if joint priors might be more suitable
* Possible (but tricky) to define all these interactions in `R-INLA` by combination of the `group` and `replicate` options
* Used sum-to-zero constraints to make posterior inferences interpretable
  * Because we're interested in the contribution of each random effect to total variance
* Model comparison via CPO statistic

## Logistic regression model

* Consider models of the form
\begin{align}
    y_{ia4}^\star &\sim \text{Binomial} \left( y_{ia3}^\star + y_{ia4}^\star, q_{ia} \right), \label{eq:logistic-regression} \\
    q_{ia} &= \text{logit}^{-1} \left( \eta_{ia} \right), \\
    \eta_{ia} &= \beta_0 +  \zeta_{c[i]} + \alpha_{ac[i]} + \phi_i + \beta_{\texttt{cfsw}} \texttt{cfsw}_{c[i]}.
\end{align}
* With the following terms:
  * $\beta_0$: intercept
  * $\zeta_{c[i]}$: country effects (IID)
  * $\alpha_{ac[i]}$ age-country effects (IID)
  * $\phi_i$: spatial effects (IID, **Besag**)
  * Clients of FSW covariates ($\texttt{cfswever}$, $\texttt{cfswrecent}$) [@hodgins2022population]

## Combination and FSW adjustment

* Take 1000 samples from each model, then multiply suitably to generate estimates for all four risk groups
* We adjusted the $k = 4$ risk group to match age-country FSW estimates
  * Obtained these by disaggregating @stevens2022estimating by age using estimates of sexually active population from @nguyen2022trends
